---
title: "Time Series Final Project Extra Analyses"
author: "Joe Coyne"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(astsa)
library(forecast)
library(imputeTS)
library(maps)
library(mapproj)
library(olsrr)
library(grid)
library(scales)
```

Explorations: - Accidents by State - Accidents by Time Zone - Accidents by Severity - Accidents by Weather - Accidents by Time of Day (Day vs Night)

```{r}
acc0 <- read.csv("US_Accidents_March23.csv")
acc0
```

```{r}
acc1 <- acc0 %>% 
  mutate(Start_Time = as.Date(Start_Time)) %>% 
  select(-End_Time, -End_Lat, -End_Lng, -Description)
acc1
```

```{r}
acc <- acc1 %>%
  group_by(Start_Time) %>%
  mutate(n = n()) %>%
  ungroup()
```

```{r}
acc_ex <- acc %>% 
  select(-ID, -Source, -Start_Lat, -Start_Lng, -Distance.mi., -Street, -City, -County, -Zipcode, -Country, -Airport_Code, -Weather_Timestamp, -Wind_Chill.F., -Humidity..., -Pressure.in., -Visibility.mi., -Wind_Direction, -Wind_Speed.mph., -Precipitation.in., -Amenity, -Bump, -Crossing, -Give_Way, -Junction, -No_Exit, -Railway, -Roundabout, -Station, -Stop, -Traffic_Calming, -Traffic_Signal, -Turning_Loop, -Civil_Twilight, -Nautical_Twilight, -Astronomical_Twilight) %>% 
  mutate(Year = year(Start_Time)) %>% 
  distinct(Start_Time, .keep_all = T)
acc_ex
```

## Accidents by State

```{r}
acc_state <- acc_ex %>% 
  select(Year, n, State) %>% 
  group_by(Year, State) %>%
  summarize(total_n = sum(n, na.rm = TRUE))
acc_state
```

```{r}
acc_state %>% 
  arrange(State)
#Some states only have data from certain years -> ex: Arkansas only has data from 2022
```

```{r}
states <- map_data("state")
```

```{r}
acc_state$State <- trimws(acc_state$State)
acc_state$state_name <- state.name[sapply(acc_state$State, \(x) which(x == state.abb)[1])]
```

```{r}
acc_state <- acc_state %>%
  mutate(region = tolower(state_name))
```

```{r}
# List of all 50 states plus Washington D.C.
all_states <- c(
  "alabama", "alaska", "arizona", "arkansas", "california", "colorado", "connecticut", 
  "delaware", "florida", "georgia", "hawaii", "idaho", "illinois", "indiana", "iowa", 
  "kansas", "kentucky", "louisiana", "maine", "maryland", "massachusetts", "michigan", 
  "minnesota", "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
  "new jersey", "new mexico", "new york", "north carolina", "north dakota", "ohio", "oklahoma", 
  "oregon", "pennsylvania", "rhode island", "south carolina", "south dakota", "tennessee", 
  "texas", "utah", "vermont", "virginia", "washington", "west virginia", "wisconsin", 
  "wyoming")

# Create a data frame with all combinations of Year and State
all_combinations <- expand.grid(
  Year = unique(acc_ex$Year), 
  region = all_states
)

# Summarize the data to get total_n by Year and State
acc_state <- acc_state %>% 
  select(Year, total_n, region) %>% 
  group_by(Year, region) %>%
  summarize(total_n = sum(total_n, na.rm = TRUE), .groups = "drop")

# Left join with the complete list of Year and State combinations
acc_state_complete <- all_combinations %>%
  left_join(acc_state, by = c("Year", "region")) %>%
  mutate(total_n = ifelse(is.na(total_n), 0, total_n))

acc_state_complete %>% 
  filter(Year == 2016)
```

```{r}
accidents.geo <- merge(states, acc_state_complete, sort = FALSE, by = "region")
accidents.geo <- accidents.geo[order(accidents.geo$order), ]
accidents.geo <- accidents.geo %>%
  mutate(fill_color = ifelse(total_n == 0, NA, total_n))  # NA for no accidents, actual total_n for non-zero
 
accidents.geo16 <- accidents.geo %>% 
  filter(Year == 2016)
accidents.geo17 <- accidents.geo %>% 
  filter(Year == 2017)
accidents.geo18 <- accidents.geo %>% 
  filter(Year == 2018)
accidents.geo19 <- accidents.geo %>% 
  filter(Year == 2019)
accidents.geo20 <- accidents.geo %>% 
  filter(Year == 2020)
accidents.geo21 <- accidents.geo %>% 
  filter(Year == 2021)
accidents.geo22 <- accidents.geo %>% 
  filter(Year == 2022)
accidents.geo23 <- accidents.geo %>% 
  filter(Year == 2023)
```


```{r}
state_totals <- acc_state_complete %>%
  group_by(region) %>%
  summarise(total_accidents = sum(total_n, na.rm = TRUE))

accidents.geo <- left_join(accidents.geo, state_totals, by = "region")

accidents.geo %>% 
  group_by(region) %>% 
  arrange(total_accidents) %>% 
  distinct(region, total_accidents)
```

```{r}
accidents_fill <- accidents.geo %>%
  group_by(region, Year) %>%
  slice(1) %>%
  ungroup()

accidents_plot <- left_join(accidents.geo, accidents_fill %>% select(region, Year, fill_color = total_accidents), 
                            by = c("region", "Year"))

ggplot(accidents_plot, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = fill_color.y), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",
    high = "blue4",
    na.value = "gray",
    name = "Total Accidents"
  ) +
  labs(title = "Total Accidents by State (2016–2023)") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```


```{r}
# Total
grand_total <- sum(acc_state_complete$total_n, na.rm = TRUE)


ggplot(accidents.geo, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = total_accidents), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State (2016-2023)") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("Grand Total:", format(grand_total, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)

# Want to add the summed total accidents for each year
```
New York and California have the most total accidents from 2016-2023 by far, with New York at 2.2 million and California at 1.9 million. Michigan (755,720), New Jersey (428,285), and Illinois (364,048) have the next highest.

Delaware, Maine, Mississippi, New Hampshire, North Dakota, Oklahoma, and South Dakota all have 0 reported total accidents. West Virginia only has 7 reported total accidents over this same time frame. These seems unlikely and are probably reporting issues.

```{r}
# 2016
grand_total16 <- accidents.geo16 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo16, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2016") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2016 Total:", format(grand_total16, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2017
grand_total17 <- accidents.geo17 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)


ggplot(accidents.geo17, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2017") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2017 Total:", format(grand_total17, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2018
grand_total18 <- accidents.geo18 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo18, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2018") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2018 Total:", format(grand_total18, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2019
grand_total19 <- accidents.geo19 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo19, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2019") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2019 Total:", format(grand_total19, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2020
grand_total20 <- accidents.geo20 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo20, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2020") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2020 Total:", format(grand_total20, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2021
grand_total21 <- accidents.geo21 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo21, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2021") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2021 Total:", format(grand_total21, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2022
grand_total22 <- accidents.geo22 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo22, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2022") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2022 Total:", format(grand_total22, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)
```

```{r}
# 2023
grand_total23 <- accidents.geo23 %>%
  group_by(region) %>%
  slice(1) %>%  # Take just one row per state
  summarise(state_total = first(total_n)) %>%
  summarise(grand_total = sum(state_total, na.rm = TRUE)) %>%
  pull(grand_total)

ggplot(accidents.geo23, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fill_color), color = "white") +
  coord_map() +
  scale_fill_gradient(
    low = "dodgerblue",  # Color for lower non-zero total_n values
    high = "blue4",  # Color for higher non-zero total_n values
    na.value = "gray",  # Color for total_n == 0 (no accidents) - will be gray
    name = "Total Accidents",
    labels = comma
  ) +
  labs(title = "Total Accidents by State - 2023") +
  theme_void() +
   theme(
    plot.title = element_text(hjust = 0.5))

grid.text(
  label = paste("2023 Total:", format(grand_total23, big.mark = ",")),
  x = unit(0.97, "npc"),  # Near right side
  y = unit(0.1, "npc"),   # Lower on the page
  just = "right",
  gp = gpar(fontsize = 10)
)

#only goes up until April 2023
```

## Regression to Predict Total Daily Accidents

```{r}
acc_ex
```

```{r}
acc_reg <- acc_ex %>% 
  mutate(State = as.factor(State), Timezone = as.factor(Timezone), Weather_Condition = as.factor(Weather_Condition), Sunrise_Sunset = as.factor(Sunrise_Sunset))

out=lm(n~Severity+State+Timezone+Temperature.F.+Weather_Condition+Sunrise_Sunset+Year, data = acc_reg)

ols_step_forward_p(out)
k=ols_step_best_subset(out)
k
```

```{r}
#summary(out)
anova(out)
```

```{r}
out2=lm(n~Severity+State+Timezone+Temperature.F.+Weather_Condition+Year, data = acc_reg)
summary(out2)
anova(out2)
```
**We find that Severity, State, Timezone, Temperature, Weather_Condition, and Year are all significant in predicting the total number of accidents in a day.**
